Abstract Spatial Specification of Stencil Computations

### 2016-06-30 paper.tex
removed INTER axiom
Proof of * commutativity and matrix notation fix
split modifieres in the grammar
butchered the abstract

\begin{abstract}
  Verifying the correctness of numerical computations is hard because
  potential specifications for their behaviour tend to be highly
  decouple from the code, defined in terms of the originating
  mathematical model. For example, a system of partial differential equations
  (PDEs) specifies a relationship between variables, but tools for encoding such
  a specification, and pertinent properties such as convergence and
  stability, are rare.  Furthermore, checking the correctness of some
  implementation of a discrete approximation against such
  specifications is extremely challenging. We seek intermediate
  systems of specification and verification between lower-level
  aspects (\eg{}, out-of-bounds errors) and full correctness
  with-respect to a numerical model. In this paper, we focus on
  \emph{stencil computations}-- a common idiom in numerical code, but
  one which is prone to error due to fine-grained programming
  mistakes.  Our conjecture is that the vast majority of stencil
  computations have a simple, regular, static shape that can be
  described by a simple specification language. We propose a language
  for this, along with an inference and checking procedure. We test
  our conjecture against a corpus of numerical Fortran code, and show
  that static specifications can be given to $X\%$ of identifiable
  stencil computations, with a mean ratio of $Y:Z$
  specification-to-program code. This paper details our stencil
  specification language, its static semantics, inference, checking,
  and program synthesis procedures, our implementation, and our
  evaluation study.
\end{abstract}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
\nonterm{spec} ::= & [\nonterm{mod}] \;
                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; &
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em]
(\textsc{inter}) \;\; & (S \; \texttt{+} \; S') \; \texttt{*} \;
                              (T \; \texttt{+} \; T') \\
 \equiv \; & (S \;  \texttt{*} \; S)
 \; \texttt{+} \; (S \; \texttt{*} \; T')
 \; \texttt{+} \; (S' \; \texttt{*} \; T)
 \; \texttt{+} \; (S' \; \texttt{*} T')  \\[0.5em]
\end{array} \\ \\
%
\framebox{${\textit{specDec} :: \texttt{v}} \equiv
{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

@@

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{align*}
M \otimes N & =
  \{(u \bowtie v)_i \mid i \in \{0, \ldots, 2^n \}, (u, v) \in
              M \times N\}
\end{align*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors).  Using this, the $\otimes$ operation constructs the set
of every pairwise permutation for every pair $(u, v)$ of $n$-vectors in the
Cartesian product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer $i$.
Recall that the vectors are defined over the domain
$\mathbb{Z} \cup \{\infty\}$.
For addition and multiplication, we take
 $\infty$ to be an absorbing element. Thus, if either
vector in a pairwise permutation contains $\infty$ in
the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
permutation is $\infty$, \eg{},
%%
\begin{equation*}
\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{cc}
0 & \infty \\
0 & \infty \\
2 & \infty \\
2 & \infty
\end{array}
\right]
\end{equation*}

>>>>

\begin{abstract}
Verifying the correctness of numerical computations relies on a specification of their behaviour.
A natural way to do this would be in terms of the original mathematical model but this approach yields specifications which are highly decoupled from the code.
Checking the correctness of code which implements a discrete approximation of such specifications is extremely challenging. 
Instead we have been investigating intermediate
  systems of specification and verification between lower-level
  aspects (\eg{}, out-of-bounds errors) and full correctness
  based on a numerical model. In this paper, we focus on
  \emph{stencil computations}---a common idiom in numerical code, but
  one which is error-prone due to fine-grained indexing errors.  
%
We describe a stencil specification language which can be directly embedded as annotations in source code, along with an inference and checking procedure. 
%
We evaluate our language against a corpus of numerical Fortran code, and show
  that precise specifications can be given to $X\%$ of identifiable
  stencil computations, with a mean ratio of $Y:Z$
  specification-to-program code and that the vast majority of stencil computations have a simple, regular, static shape.
%
This paper details our stencil
  specification language, its static semantics, inference, checking,
  and program synthesis procedures, our implementation, and our
  evaluation study.
%
We provide an implementation of stencil inference and validation in \emph{Camfort}, an open-source program analysis and refactoring tool.
\end{abstract}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{mod}] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; &
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em]
\end{array} \\ \\
%
\framebox{${\textit{specDec} :: \texttt{v}} \equiv
{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

@@

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{align*}
M \otimes N & =
  \{(u \bowtie v)_i \mid i \in \{1, \ldots, 2^n\}, (u, v) \in M \times N\}
\end{align*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain
$\mathbb{Z} \cup \{\infty\}$.
For addition and multiplication, we take
 $\infty$ to be an absorbing element. Thus, if either
vector in a pairwise permutation contains $\infty$ in
the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
permutation is $\infty$, \eg{},
%%
\begin{equation*}
\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{cc}
0 & \infty \\
0 & \infty \\
2 & \infty \\
2 & \infty
\end{array}
\right]
\end{equation*}
